---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ellipsoidgaussian

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/ellipsoidgaussian)](https://CRAN.R-project.org/package=ellipsoidgaussian)
[![R-CMD-check](https://github.com/hanyu-song/ellipsoidgaussian/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hanyu-song/ellipsoidgaussian/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Description 
The package provides Bayesian estimation of the ellipsoid-Gaussian model
    introduced by @song2022curved. The model was inspired by the
    need for new models for characterizing dependence in multivariate
    data. The multivariate Gaussian distribution is routinely used, but
    cannot characterize nonlinear relationships in the data. Most
    non-linear extensions tend to be highly complex; for example,
    involving estimation of a non-linear regression model in latent
    variables. The relatively simple class of ellipsoid-Gaussian
    multivariate distributions, which are derived by using a Gaussian
    linear factor model involving latent variables having a von
    Mises-Fisher distribution on a unit hyper-sphere, can be used for
    modeling data distributed around a hyper-ellipsoid. In addition to
    fitting the model, the package also provides options for
    post-processing the posterior samples and generating samples from the
    posterior predictive distribution. Plotting functions are also
    included.

## Installation

You can install the development version of ellipsoidgaussian from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("hanyu-song/ellipsoidgaussian")
```

The package uses Cayley transform ellipsoid fitting [@melikechi2023ellipsoid]
for initial values of the parameters in the model, which depends on 
[the Python module ctef](https://github.com/omelikechi/ctef). The module is NOT 
automatically installed when the package is installed. To install the module,
follow the instruction on the [GitHub page](https://github.com/omelikechi/ctef).
Make sure the module is installed in the path as the Python used in R; run
``` r
reticulate::py_config()
```
to see the path to Python.

## Example

```{r example}
library(ellipsoidgaussian)
```
Here we demonstrate the usage of the package using the included horse mussels 
data; more details can be found in the vignette.

Use the help function to learn about the data set:

```{r help}
help(mussels)
```


To visualize the data, do 

```{r visualize}
pairsPlot_onedata(scale(mussels))
```

To fit an ellipsoid-Gaussian model, run
```{r fit the model}
samples <- ellipsoid_gaussian(dat = mussels,
                                k = 2, 
                                scale_col = TRUE, 
                                updateCenter = TRUE, 
                                niter = 5000)
```

Then we can generate posterior predictive samples using
```{r generate posterior predictive}
  ppd <- gen_posterior_predictive(n = 200, samples = samples, burnin = 2500)
```

We can also visualize the posterior predictive samples in juxatoposition with the original data using

```{r visualize results}
  pairsPlot(scale(mussels), ppd)
```


Factor loadings are unidentifiable in a factor model. To post process the samples, 
run
```{r post process the factor loadings}
  aligned <- postprocess(scale(mussels), samples, burnin = 2500)

```

After post-processing the samples, visualize the posterior mean of the factor loadings for interpretation.

```{r plot the factor loadings}
  plot_factor_loadings(lambda_samples = aligned$lambda, row_idx = 1:4) 
```
This suggests that the shell size, as measured by an average of height, length and width, is a good indicator of the muscle mass. 

If you find bugs or want to suggest new features please visit the
**ellipsoidgaussian** [GitHub issues
page](https://github.com/hanyu-song/ellipsoidgaussian/issues).

## References

