---
title: "The Ellipsoid-Gaussian model"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{The Ellipsoid-Gaussian model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

**ellipsoidgaussian** is an R package that provides an efficient
function for fitting the ellipsoid-Gaussian model using a Bayesian
method. The model, introduced by @song2022curved, was inspired by the
need for new models for characterizing dependence in multivariate data.
The multivariate Gaussian distribution is routinely used, but cannot
characterize nonlinear relationships in the data. Most non-linear
extensions tend to be highly complex; for example, involving estimation
of a non-linear regression model in latent variables.

The relatively simple class of ellipsoid-Gaussian multivariate
distributions, which are derived by using a Gaussian linear factor model
involving latent variables having a von Mises-Fisher distribution on a
unit hyper-sphere. Mathematically, it is defined as
$$  {x}_i = {c} + {\Lambda} {\eta}_i + \epsilon_i, \quad \epsilon_i \sim \text{N}_p(0, \Sigma), \quad \eta_i \sim \text{vMF}(\mu, \tau),\quad i=1,\ldots,n,\label{equ:egdefinition}$$
where $i$ denotes the observation index for the $n$ observations in the
data set, $x_i, \eta_i$ and $\epsilon_i$ represent the observation
vector, the latent variable and noise vector associated with observation
$i$. $c, \Lambda, \Sigma := \text{diag}(\sigma_1^2, \ldots, \sigma_p^2), \mu$ and $\tau$ are the parameters for the
model. Integrating over the distribution of the latent factors yields
the ellipsoid-Gaussian (marginal) distribution. The density is
$$f_\text{EG}(x) = \frac{C_k(\tau)}{(2\pi)^{\frac{p}{2}}\prod_{i = 1}^p \sigma_i} \exp\left\{-\frac{1}{2} (x - {c})^T \Sigma^{-1} (x - {c})\right\}
   \varsigma\left\{ \tau \mu + \Lambda^T \Sigma^{-1
}(x - {c}) , \frac{\Lambda^T \Sigma^{-1}\Lambda}{2}\right\},$$ with
respect to the Lebesgue measure on $\mathbb{R}^p,$ where
$\varsigma(\kappa \vartheta, A)$ is the normalizing constant in a
Fisher-Bingham density that is defined as
$$f({y}; \kappa, \vartheta, A) = \frac{1}{\varsigma(\kappa \vartheta, A)} \exp\left(\kappa \vartheta^T{y} - {y}^T A {y}\right),$$
with respect to $\mathscr{S}^{k-1},$where ${y}, \vartheta \in \mathbb{S}^{k -1}$, $\kappa \geq 0$, $A \in \mathbb{R}^{p \times p}$ is a symmetric matrix, and $\varsigma(\kappa \vartheta, A)$ is a normalizing constant that
is best numerically approximated via saddle point approximation [@KumeWood05].

Ellipsoid-Gaussian distribution can flexibly model
curved relationships among variables with lower-dimensional structures; see @song2022curved for more details.

#### Installing ellipsoidgaussian 
**ellipsoidgaussian** will run in Windows, Mac OS X, or Linux. To install **ellipsoidgaussian** you first need to install R and we highly recommend install RStudio as well. 

Type in R 
```{r, eval = FALSE}
install.packages('devtools')
devtools::install_github()
```
to install the latest development version of the package from the Github page.

To instead install the latest official release of the package from CRAN, type
```{r, eval = FALSE}
install.packages('ellipsoidgaussian')
```

In either case, if you type
```{r set up}
library(ellipsoidgaussian)
```

it will load all the **ellipsoidgaussian** functions. Note that Python and the `ctef` module installation (CHANGE: link) is required for `gen_input_eg()`. To check whether the module is successfully installed, run the following command:
```{r}
reticulate::py_module_available('ctef')
```

One can simulate data from an ellipsoid-Gaussian distribution using the `rellipsoidgaussian()` function, specifying the sample size n and the parameters $c$ (center), $\Lambda$ (Lambda), $\mu$ (mu), $\tau$ (tau) and $\Sigma$ (Sigma), e.g.
```{r generate from EG}
  dat <- rellipsoidgaussian(n = 1000, 
                          center = rep(0,3), 
                        Lambda = matrix(rnorm(6, mean = -5, sd = 2),nrow = 3, ncol = 2),
                          mu = c(0,1), 
                          tau = 1, 
                          Sigma = diag(0.5,3))

```

The shell data included in the package is an example data set simulated from the model. In this vignette, we use the horse mussels data. You can load the data via
```{r load data}
data(mussels)
```

Use 
```{r help data}
help(mussels)
```
to learn about the data set. One interest of the study was to understand the relationship of muscle mass, the mass of the edible portion, to the measurements of the horse mussels. In the analysis below, we scale the data since each measurement is on different scales.

##### Visualizing the data 

The package supports visualization using scatter plot matrices, with correlation in the upper triangular part, density on the diagonal, and scatter plots on the lower triangular part. 

Before fitting the model, we can visualize the data to detect the curvature.

```{r visualize}
pairsPlot_onedata(scale(mussels))
```

###### Fitting the model & running the MCMC chain

Taking a Bayesian approach, we propose a hybrid of geodesic stochastic gradient Nosé-Hoover Thermostats (gSGNHT) [@liu16sgmcmc] and adaptive Metropolis [@vihola12AM] for posterior sampling. The prior distribution is specified on the help page of `gen_input_eg()`. The initial values of the parameters are produced by `gen_input_eg()` using  @melikechi2023ellipsoid. `ellipsoid_gaussian()` automates the model fitting process by first generating the initial values and then sampling from the posterior distribution. It also allows the option to update or not update the center $c$ during the sampling process. When the center is not updated, it is fixed at the initial value. Users can also specify the algorithm-specific parameters, including the minibatch size,the step size, and the scalar parameter, all of which are tuning parameters for the gSGNHT algorithm; see @liu16sgmcmc for more details. The default values for these parameters are 50, $10^{-4}$ and 0.1. Other parameters that require user input are `k` (The number of latent dimensions), `scale_col` (Whether each variable should be scaled to have mean 0 and variance 1) and `niter` (total number of iterations for the sampling process). The output are saved in a list, with samples associated with each parameter saved as an element in the list.

```{r fit the model}
if (reticulate::py_module_available('ctef')) {
  samples <- ellipsoid_gaussian(dat = mussels,
                                k = 3, 
                                scale_col = FALSE, 
                                updateCenter = TRUE, 
                                niter = 5000)
}
```

##### Generating posterior predictive samples
We can generate posterior predictive samples using the output from ``ellipsoid_gaussian``.

```{r generate posterior predictive}
if (reticulate::py_module_available('ctef')) {
  ppd <- gen_posterior_predictive(n = 200, samples = samples, burnin = 2500)
}
```

##### Visualizing ellipsoid-Gaussian results
We can both visualize the posterior predictive samples alone and visualize the samples with the original data set in juxtaposition. To visualize just the posterior predictive samples, we run
```{r visualize result 1}
if (reticulate::py_module_available('ctef')) {
  pairsPlot_onedata(ppd)
}
```
To visualize the two data sets in juxtaposition, we run
```{r visualize results}
if (reticulate::py_module_available('ctef')) {
  pairsPlot(scale(mussels), ppd)
} 
```

##### Postprocessing the factor loadings 

Factor loadings are unidentifiable in a factor model. We perform the varimax rotation on the factor loadings samples to resolve rotational ambiguity. See 

```{r}
help(jointRot, package = "infinitefactor")
```

for more information. Run `postprocess()` for post processing of the factor loadings and the latent factors jointly. 
```{r post process the factor loadings}
if (reticulate::py_module_available('ctef')) {
  aligned <- postprocess(scale(mussels), samples, burnin = 2500)
}
```

Then we visualize the posterior mean of the post-processed factor loadings for interpretation.

```{r plot the factor loadings}
if (reticulate::py_module_available('ctef')) {
  p1 <- plot_factor_loadings(lambda_samples = aligned$lambda, row_idx = 1:4) 
  print(p1)
}

```

This suggests that the shell size, as measured by an average of height, length and width, is a good indicator of the muscle mass. 

If you find bugs or want to suggest new features please visit the
**ellipsoidgaussian** [GitHub issues
page](https://github.com/hanyu-song/ellipsoidgaussian/issues).

##### References
